WASP-4\lowercase{b} Arrived Early for the TESS Mission

### 2018-12-11 paper/ms.tex
update ms.pdf

\begin{abstract}
  %TODO: final numbers
  By combining data from the Transiting Exoplanet Survey Satellite (TESS)
  with data from previous studies, we show that the transit times of WASP-4b are
  incompatible with a constant orbital period.  The
  transits observed by TESS occurred $77.8 \pm 10.8$~seconds earlier than one would expect, based on the previous data.
  The orbital period appears to be shrinking by $\dot{P}=-12.1 \pm 1.2$
  milliseconds per year.   A simple clock offset is unlikely, because TESS observations of other
  hot Jupiter (WASP-6b, 18b, and 46b) are compatible with a constant period and rule out
  a 77.8-second offset at the 6.3$\sigma$ level.
  The timing variations seem to be astrophysical,
  and could be caused by apsidal precession, or perhaps by tidal
  decay.  The latter seems implausible on theoretical grounds, as it
  would require a stellar quality factor $Q_\star' \approx
  3\times10^4$.  Apsidal precession would imply an eccentricity
  $e\approx1.6\times 10^{-3}$, and the possibility of measuring the
  planet's Love number.  A Doppler shift from WASP-4's acceleration
  towards us could account for at most one quarter of the observed
  period change (at $2\sigma$).  Further observations will help
  confirm the reality of the timing variation, and eventually
  determine its cause.
\end{abstract}

@@

Using the cleaned PDC lightcurve, we ran Box Least Squares (BLS) to
estimate the transit epoch, period, and duration
\citep{kovacs_box-fitting_2002}.  Using these parameters, we isolated
each transit to within $\pm$10 transit durations of its midtime.  In
each transit window, we then simultaneously fit for a model transit,
plus a local linear trend.  Our transit model uses the analytic
formulae calculated by \citet{mandel_analytic_2002} and implemented by
\citet{kreidberg_batman_2015}.

We allowed four free parameters per transit: the time of mid-transit
$t_{\rm tra}$, the planet-to-star radius ratio $R_{\rm p}/R_\star$,
and the slope and intercept of the line.
% Our priors were wide uniform distributions centered on the BLS
% estimates for $t_{\rm tra}$ and $R_{\rm p}/R_\star$, and on a slope
% of zero and an intercept of unity for the line.
We fixed the remaining transit parameters as follows.  We set the
eccentricity to zero, and the longitude of periastron to $\pi/2$.  We
assumed a quadratic limb-darkening law, with coefficients interpolated
from the \citet{claret_limb_2017} tables.  For the period,
$a/R_\star$, and inclination, we adopted the values
from~\citet{petrucci_no_2013}.  These are in reasonable agreement with
the parameters reported by \citet{gillon_improved_2009},
\citet{southworth_high-precision_2009}, and
\citet{huitson_gemini_2017}.

Having defined the free and fixed parameters, we calculated an initial
guess for the free parameters by maximizing the likelihood.  We then
sampled over the posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013}.
% During this process, we used the \texttt{corner} software to check
% that our posteriors were being well-sampled~\citep{corner_2016}.
After completing an initial fit, we set the photometric error bars to
be equal to the standard deviation of the in-transit points of the
residual lightcurve.  We then reperformed the fitting procedure, using
the empirically determined photometric errors.

@@

Figure~\ref{fig:lightcurves} shows the light curves, best-fit models,
and residuals.  Table~1 reports the mid-transit times and their
uncertainties.  Taking the mean and standard deviation of the measured
planet to star radius ratio, we find $R_{\rm
p}/R_\star=0.1538\pm0.0013$, in agreement with previous determinations
\citep{wilson_wasp-4b_2008,gillon_improved_2009,winn_transit_2009,southworth_high-precision_2009}.
Binning the residuals to 1-hour windows and taking the median absolute
deviation, the lightcurves have a MAD of $594\,{\rm ppm}\,{\rm
hr}^{1/2}$.  The predicted\footnote{\url{github.com/lgbouma/tnm}}
photon-counting noise is $410\,{\rm ppm}\,{\rm hr}^{1/2}$.  At
$T\approx11.8$, with a 9-pixel aperture read noise is expected to
contribute an additional $202\,{\rm ppm}\,{\rm hr}^{1/2}$ in
quadrature.  The predicted zodiacal background contribution is
$673\,{\rm ppm}\,{\rm hr}^{1/2}$, based on the
\citet{winn_photonflux_2013} model, which \citet{Sullivan_2015} used
in planet detection simulations.  This zodiacal noise prediction seems
to have been an overestimate, given the observed MAD of WASP-4.  A
more detailed assessment of TESS's photometric performance is beyond
the scope of this work.

@@

We used the homogeneous \citet{hoyer_tramos_2013} timing study as our
starting point.  Though their latest two epochs fell slightly below
the prediction of a linear ephemeris (Panel C of their Figure~6), they
did not find a need to fit a quadratic function to the O-C values.  We
verified that all times were on the BJD$_{\rm TDB}$ time standard
using the \citet{eastman_achieving_2010} calculator.  We set the
zero-point epoch to be as close as possible to the average of the
midtimes, weighted as the inverse midtime variance. This minimizes the
covariance between the transit epoch and the period
\citep{gibson_gemini_2013}.

>>>>

\begin{abstract}
  %TODO: final numbers
  By combining data from the Transiting Exoplanet Survey Satellite
  (TESS) with data from previous studies, we show that the transit
  times of WASP-4b are incompatible with a constant orbital period.
  The transits observed by TESS occurred $77.8 \pm 10.8$~seconds
  earlier than one would expect, based on the previous data.  The
  orbital period appears to be shrinking by $\dot{P}=-12.1 \pm 1.2$
  milliseconds per year.   A simple clock offset is unlikely, because
  TESS observations of other hot Jupiters (WASP-6b, 18b, and 46b) are
  compatible with a constant period and rule out a 77.8-second offset
  at the 6.3$\sigma$ level.  The timing variations seem to be
  astrophysical, and could be caused by apsidal precession, or perhaps
  by tidal decay.  The latter seems implausible on theoretical
  grounds, as it would require a stellar quality factor $Q_\star'
  \approx 3\times10^4$.  Apsidal precession would imply an
  eccentricity $e\approx1.6\times 10^{-3}$, and the possibility of
  measuring the planet's Love number.  A Doppler shift from WASP-4's
  acceleration towards us could account for at most one quarter of the
  observed period change (at $2\sigma$).  Further observations will
  help confirm the reality of the timing variation, and eventually
  determine its cause.
\end{abstract}

@@

Using the cleaned PDC lightcurve, we ran Box Least Squares (BLS) to
estimate the transit epoch, period, and duration
\citep{kovacs_box-fitting_2002}.  Using these parameters, we isolated
each transit to within $\pm$5 transit durations of its midtime.  In
each transit window, we then simultaneously fit for a model transit,
plus a local linear trend.  Our transit model uses the analytic
formulae calculated by \citet{mandel_analytic_2002} and implemented by
\citet{kreidberg_batman_2015}.

We allowed four free parameters per transit: the time of mid-transit
$t_{\rm tra}$, the planet-to-star radius ratio $R_{\rm p}/R_\star$,
and the slope and intercept of the line.
% Our priors were wide uniform distributions centered on the BLS
% estimates for $t_{\rm tra}$ and $R_{\rm p}/R_\star$, and on a slope
% of zero and an intercept of unity for the line.
We fixed the remaining transit parameters as follows.  We set the
eccentricity to zero. We
assumed a quadratic limb-darkening law, with coefficients interpolated
from the \citet{claret_limb_2017} tables.  For the period,
$a/R_\star$, and inclination, we adopted the values
from~\citet{petrucci_no_2013}.  These are in reasonable agreement with
the parameters reported by \citet{gillon_improved_2009},
\citet{southworth_high-precision_2009}, and
\citet{huitson_gemini_2017}.

Having defined the free and fixed parameters, we calculated an initial
guess for the free parameters by maximizing the likelihood.  We then
sampled over the posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013}.
% During this process, we used the \texttt{corner} software to check
% that our posteriors were being well-sampled~\citep{corner_2016}.
After completing an initial fit, we set the photometric error bars to
be equal to the standard deviation of the out-of-transit points of the
residual lightcurve.  We then reperformed the fitting procedure, using
the empirically determined photometric errors.

@@

Figure~\ref{fig:lightcurves} shows the light curves, best-fit models,
and residuals.  Table~1 reports the mid-transit times and their
uncertainties.  Taking the mean and standard deviation of the measured
planet to star radius ratio, we find $R_{\rm
p}/R_\star=0.1538\pm0.0013$, in agreement with previous determinations
\citep{wilson_wasp-4b_2008,gillon_improved_2009,winn_transit_2009,southworth_high-precision_2009}.
Binning the residuals to 1-hour windows and taking the median absolute
deviation, the lightcurves have an RMS of $586\,{\rm ppm}\,{\rm
hr}^{1/2}$.  The predicted\footnote{\url{github.com/lgbouma/tnm}}
photon-counting noise is $410\,{\rm ppm}\,{\rm hr}^{1/2}$.  At
$T\approx11.8$, with a 9-pixel aperture read noise is expected to
contribute an additional $202\,{\rm ppm}\,{\rm hr}^{1/2}$ in
quadrature.  The predicted zodiacal background contribution is
$673\,{\rm ppm}\,{\rm hr}^{1/2}$, based on the
\citet{winn_photonflux_2013} model, which \citet{Sullivan_2015} used
in planet detection simulations.  This zodiacal noise prediction seems
to have been an overestimate, given the observed RMS of WASP-4.  A
more detailed assessment of TESS's photometric performance is beyond
the scope of this work.

@@

We used the homogeneous \citet{hoyer_tramos_2013} timing study as our
starting point\footnote{We also tried using times directly from the
original authors; the results did not change, so we opted for the
\citet{hoyer_tramos_2013} times as the more homogeneous option.}.
Though their latest two epochs fell slightly below the prediction of a
linear ephemeris (Panel C of their Figure~6), they did not find a need
to fit a quadratic function to the O-C values.  We verified that all
times were on the BJD$_{\rm TDB}$ time standard using the
\citet{eastman_achieving_2010} calculator.  We set the zero-point
epoch to be as close as possible to the average of the midtimes,
weighted as the inverse midtime variance. This minimizes the
covariance between the transit epoch and the period
\citep{gibson_gemini_2013}.
